---
title: "America's College Promise Plan"
author: "Will Doyle"
date: "9/13/2021"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r,echo=FALSE,message=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning=FALSE, results ='hide',include=TRUE,message=FALSE)
library(tidyverse)
library(writexl)
library(scales)
```

```{r}
addir<-"../data/cleaned"
outdir<-"../output"
dir.create(outdir, showWarnings = FALSE)
```

```{r}
inst<-read_csv(file.path(addir,"institutions.csv"))
```


The Biden plan for free community college represents a huge change in
how the federal government funds higher education. For the first time,
the federal government will directly fund states to cover the full cost
of tuition for students, using a large scale federal-state matching plan
that is similar in some ways to how the federal government funds other
priorities like health care or transportation.

Over the last few days I've been trying to answer the following
questions:

Which institutions in each state will be eligible? How much will the
plan provide? How much will states have to pay? What will be the
effective subsidy among all college students by state? Which states will
see increases or decreases in per-student funding levels under the
provision of the plan? Which states will gain or lose revenues based on
participating in the plan.

Below are my current provisional answers to those questions.

The basic concept of the plan is this. The federal government will
provide a subsidy to each state equal to eighty percent of\
(a sort of) average of nationwide community college tuition times the
number of full time equivalent students enrolled in community colleges.
States in return must reduce tuition to 0 for all students enrolled in
community colleges for their first sixty credits (two years, assuming 15
credits/semester) and must commit to providing a matching set of funds,
topping out at 20 percent after a few years.

In this writeup I describe the implications for states of the Biden free
community college plan. Which states will receive more funding, which
states will receive less, and why?

# Defining Community Colleges

The first question is: which institutions (and then students) are
eligible?

As Kevin Carey and others have noted, there's no singular definition of
a community college across the states. In fact, there are states do not
have any institutions named "community colleges" Here's how the plan
defines community colleges:

"(2) COMMUNITY COLLEGE.---The term 'community college' means--- a
degree-granting public institution of higher education at which--- the
highest degree awarded is an associate degree; or an associate degree is
the predominant degree awarded; "

Using data from the federal government's IPEDS database of colleges and
universities, I used this definition to create a list of all
institutions in every state that would qualify as community colleges.

```{r}
cc_inst<-inst%>%
  group_by(stabbr)%>%
  mutate(state_fte=sum(fteug,na.rm=TRUE))%>%
  ungroup()%>%
  rowwise()%>%
  mutate(most_degrees=max(Bachelors, ## Select the largest from this list
                          Masters, 
                          PhD,
                          Associates))%>%
  mutate(predom_assoc=ifelse(most_degrees==Associates,1,0))%>% ## If the larges is associates then the inst is predom associ
  mutate(promise_elig=ifelse(predom_assoc==1|hloffer==3,1,0 ))%>% ## OR highest offer is associates
  mutate(total_degree=rowSums(across(.cols=c("Bachelors","Masters","PhD","Associates")),na.rm=TRUE))%>%
  mutate(prop_assoc=Associates/total_degree)%>%
  filter(control==1)%>% ## public only
  filter(promise_elig==1) ## eigible only
  
sheet_list<-list()
for (st in unique(cc_inst$stabbr)){
cc_inst_sub<-cc_inst%>%filter(stabbr==st)%>%select(stabbr,instnm,control,sector,carnegie,prop_assoc)
sheet_list[[st]]<-cc_inst_sub

 }
names(sheet_list)<-unique(cc_inst$stabbr) 



write_xlsx(sheet_list,path = file.path(outdir,"eligible_insts.xlsx") )      
```

There are about `r dim(cc_inst)[2]` institutions eligible for the plan, using these
criteria. The list is available in [excel format](https://github.com/wdoyle42/cc_promise_plan/raw/main/output/%20eligible_insts.xlsx)

In most states institutions award either associates or bachelors
degrees. In a few states there are a number of institutions that are
fairly close to this cutoff. For instance, Albany State and College of
Coastal GA are eligible, but just barely.

## Eligibile Institutions in Georgia

```{r,results='asis'}
cc_inst%>%filter(stabbr=="GA")%>%select(stabbr,instnm,control,sector,carnegie,prop_assoc)%>%arrange(-prop_assoc)%>%knitr::kable()
```

The big question about eligibility moving forward will be whether
institutions that are currently Bachelor's degree granting begin
awarding associates degrees in passing. For a substantial number of
institutions this could tip the balance to becoming predominantly
associates degree granting.

## Median Community College Tuition Across the States

The per-student subsidy for the plan will be based on an overall average
of tuition across the states, specifically:

"The median resident community college tuition and fees per student in
all States, not weighted for enrollment, for the most recent award year
for which data are available; "

```{r,results='asis'}
ak_avg<-inst%>%
  filter(stabbr=="AK",control==1)%>%
  summarize(ak_mean_tution=mean((tuition2+fee2),na.rm=TRUE))%>%
  as_vector()

cc_tuit_table<-cc_inst%>%
  group_by(name)%>%
  summarize(mean_tuition_fees=mean((tuition2+fee2),na.rm=TRUE))%>%
  mutate(mean_tuition_fees=ifelse(name=="Alaska",ak_avg,mean_tuition_fees))%>%
  arrange(-mean_tuition_fees)%>%
    mutate(mean_tuition_fees=dollar(mean_tuition_fees))%>%
  rename("Average State Tuition and Fees"=mean_tuition_fees)
  knitr::kable(cc_tuit_table)

median_tuition=cc_inst%>%
  group_by(name)%>%
  summarize(mean_tuition_fees=mean((tuition2+fee2),na.rm=TRUE))%>%
  mutate(mean_tuition_fees=ifelse(name=="Alaska",ak_avg,mean_tuition_fees))%>%
  summarize(median_tuition_fees=median(mean_tuition_fees))%>%
  as_vector()


fed_subsidy=.8*median_tuition

state_match=.2*median_tuition

```



Given that this says all states I'm going to assume that they mean the
median of all *average* state tuition and required fees. That is, I take the average
unweighted tuition and fees in each state, then take the median of that measure
for all states. It looks like that's Oklahoma and West Virginia , with an average in
state CC tuition and fees of `r median_tuition`. Part of this depends on how Alaska is
handled, but something around \$4,600 seems most likely. This will of course depend on what happens to community college tuitions in the next couple of years.  

## How much will states get?

When fully phased in, states will receive `r fed_subsidy` per full time
equivalent student in community colleges enrolled in the first six
semesters of enrollment. (Question-- is this JUST enrollment at the one
community college or first sixe semesters overall?)

The total amount that states receive isn't a super interesting question.
Big states will get a lot, small states not so much. Right now the best
way I can think of to normalize this measure is to compare it to ALL fte
in the state. Another comparison would be to all 18-24 year olds in the
state.

```{r}
cc_inst<-cc_inst%>%
  mutate(fed_spend=(median_tuition*.8)*fteug)%>%
    mutate(state_match=(median_tuition*.2)*fteug)
```

The subsidy per ALL college students will depend crucially on the
distribution of FTE enrollment by state. States with a large proportion
of students in eligible institutions will get more, states with more
four year enrollment, less.

## Total Federal Spending for ACP by Total (not CC) FTE Enrollment

```{r,fig.height=10,fig.width=7}
cc_inst%>%
  group_by(name)%>%
  summarize(fed_spending=sum(fed_spend),state_fte=mean(state_fte))%>%
  mutate(fed_spending_fte=fed_spending/state_fte)%>%
  mutate(name=fct_reorder(name,fed_spending_fte))%>%
  ggplot(aes(x=name,y=fed_spending_fte))+
  geom_col(fill="lightblue")+
  coord_flip()+
  theme_minimal()+
  xlab("")+
  ylab("Federal Spending per All FTE in State")
```

Subsidies are just basically a function of community college enrollment.
So, the big winners will be Wyoming, New Mexico and California. Vermont
and New Hampshire, not so much. Some big states will get pretty low
per-student subsidies, notably New York, Pennsylvania and Massachusetts.

A big second order effect would be the shift in enrollment to community
colleges. In states like TN that implemented some version of free or
reduced price community college for a set of students, there was a
fairly substantial shift away from four-year institutions toward
community colleges. This would of course change the overall subsidy
pattern substantially.

## What will this do to tuition+state+local approps if implemented?

Here's the basic formula for this plan again:

-   States give up tuition revenue for community college students;
-   States get 80 percent of nationwide median tuition for each fte
    enrolled in community college,
-   States must increase spending by 20% of the total sum spent in state
    (when fully phased in).

So OLD total revenues consisted of state appropriations plus local
appropriations plus tuition revenues.

NEW total revenues will consist of state appropriations plus local
appropriations plus federal funding (`r fed_subsidy` per student) plus
state match (`r state_match`) per student once fully funded). There's
a maintenance of effort provision so states are mandated not to decrease
funding.

```{r}
cc_totals<-cc_inst%>%
  mutate(eg_revs=tuition_revs+
           state_approps+
           local_approps)%>%
  mutate(state_local_revs=state_approps+local_approps)%>%
  mutate(new_eg_revs=state_approps+local_approps+fed_spend+state_match)%>%
  group_by(name)%>%
  summarize(total_revs=sum(eg_revs,na.rm=TRUE),
            total_new_revs=sum(new_eg_revs,na.rm=TRUE),
            total_state_local=sum(state_local_revs,na.rm = TRUE),
            total_fte=sum(fteug))%>%
  mutate(eg_revs_fte=total_revs/total_fte)%>%
  mutate(state_local_fte=total_state_local/total_fte)%>%
  mutate(new_eg_revs_fte=total_new_revs/total_fte)%>%
  mutate(diff=new_eg_revs_fte-eg_revs_fte)%>%
  arrange(-diff)
```

The following table shows current education and general revenues (state
and local appropriations plus tuition) per full time equivalent
community college student in each states, followed by the amount of
revenues from the proposal-- state and local appropriations plus federal
funding plus state match.

```{r,results='asis'}

cc_totals_sum<-cc_totals%>%
  filter(!name=="Alaska")%>%
  select(name,eg_revs_fte,new_eg_revs_fte,diff)%>%
  rename(State=name,"Current EG Revenues"=eg_revs_fte,"Proposed EG Revenues"=new_eg_revs_fte,Difference=diff)
  cc_totals_sum%>%knitr::kable()
```

It looks like California community college students stand to gain
\$2,800 each, while Vermont students would see a decline in per student
funding of about \$5,800 unless the state steps in and makes up the
difference. There are 14 states where the federal plus state match would
be lower than current tuition revenues.

Here's a plot showing old and new revenues by state:

```{r, fig.height=8}
cc_totals%>%
  filter(!name=="Alaska")%>%
  select(name,eg_revs_fte,new_eg_revs_fte)%>%
  mutate(name=fct_reorder(name,new_eg_revs_fte))%>%
  rename(State=name,"Current EG Revenues"=eg_revs_fte,"Proposed EG Revenues"=new_eg_revs_fte)%>%
  pivot_longer(cols=-State,names_to="Funding Type")%>%
  ggplot(aes(x=State,y=value,fill=`Funding Type`))+
  geom_bar(stat="identity",position="dodge")+
  coord_flip()+
  theme_minimal()+
  theme(legend.position = "bottom")
```

Here's a more detailed table with all of the components that go into
this calculation.

```{r,results='asis'}
spend_sum<-cc_inst%>%
  filter(!stabbr=="AK")%>%
  group_by(name)%>%
  summarize(across(.cols=c("fteug","tuition_revs","state_approps","local_approps","fed_spend","state_match"),.fns=sum,na.rm=TRUE ))%>%
  mutate(across(c("tuition_revs","state_approps","local_approps","fed_spend","state_match"), .fns= ~.x/fteug  ))%>%
  mutate(new_total=state_approps+local_approps+fed_spend+state_match)%>%
  mutate(old_total=state_approps+local_approps+tuition_revs)%>%
  mutate(name=fct_reorder(name,new_total))
knitr::kable(spend_sum)
```

And here's the likely distribution of revenues across the major sources
by state:

```{r,fig.height=10,fig.width=7}
spend_sum%>%
  select(name,state_approps,local_approps,fed_spend,state_match)%>%
  rename("Federal Spending"=fed_spend,"Local Approps"=local_approps,"State Approps"=state_approps,"State Match"=state_match)%>%
  pivot_longer(cols=-name,names_to = "Type")%>%
  mutate(Type=fct_relevel(Type,c("State Match","Federal Spending","State Approps","Local Approps")))%>%
  ggplot(aes(y=value,x=name,fill=Type))+
  geom_col()+
  coord_flip()+
    theme_minimal()+
  theme(legend.position = "bottom")+
  xlab("")

```

From the state perspective, one way to look at this is just the simple
tradeoff between new revenues from the feds minus lost revenues from
tuition plus extra revenues from the state. If this is positive,then the
state will be getting more from the feds in the subsidy than it will be
losing from the tuition revenue it's foregoing plus the state match it
will have to spend.

```{r,results='asis'}
spend_sum<-cc_inst%>%
  group_by(name)%>%
  summarize(across(.cols=c("fteug","tuition_revs","state_approps","local_approps","fed_spend","state_match"),.fns=sum,na.rm=TRUE ))%>%
  mutate(across(c("tuition_revs","state_approps","local_approps","fed_spend","state_match"), .fns= ~.x/fteug  ))%>%
  mutate(state_match_plus_tuition_fee_revs=tuition_revs+state_match)%>%
  mutate(diff=fed_spend-state_match_plus_tuition_fee_revs)%>%
  arrange(-diff)%>%
  mutate(name=fct_reorder(name,diff))%>%
  ungroup()%>%
  select(name,fed_spend,state_match_plus_tuition_fee_revs,diff)
  knitr::kable(spend_sum)
```

From this perspective, it looks like there are about
`r length(spend_sum$diff[spend_sum$diff<=0])` 
states that would have to
increase funding for community college students in order to participate.
Of these,
`r  length(spend_sum$diff[spend_sum$diff<0&spend_sum$diff>=-500])` would
have to increase funding by
\$500 or so, 
`r length(spend_sum$diff[spend_sum$diff<(-500)&spend_sum$diff>=-1000])` 
would have to increase funding by \$500-1000,

`r length(spend_sum$diff[spend_sum$diff<(-1000)&spend_sum$diff>=-2000])` 
would have to increase funding by \$1000-2000 and 
`r length(spend_sum$diff[spend_sum$diff<=-2000])`
would have to increase funding by more than \$2000 if they wanted
per-student subsidy levels at community colleges to remain the same.

Like I said, not sure if this is correct yet, I welcome comments and
criticisms: what am I getting wrong? Code and data available here for
anyone else who wants to use it.


2021-09-14: Updates! 

In a previous version I did not include required fees in calculating median tuition and fees across the states. This has been updated above. 

Similarly the list of institutions included an error, and the correct number is now included. 


